---
title: "Getting Started with ecem"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with ecem}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What this vignette is

**ecem** replaces a single fixed coarsening choice in coarsened exact
matching (CEM) with an elicited *prior* over cutpoints, draws repeatedly
from it, matches and estimates on each draw, and pools the results --
along with diagnostics for when that pooling is (and is not)
trustworthy. The methodology is described in "Elicited Priors for
Coarsening Regimes in Coarsened Exact Matching"; this vignette does not
re-derive it, but walks through the package's workflow end to end on
simulated data, in enough detail to use as a template for your own
replication analysis.

Three things this vignette covers, in order:

1. **One elicited covariate, one theory of where to coarsen it** ("the
   one-regime case") -- the basic draw-match-pool workflow.
2. **One elicited covariate, two competing theories of where to coarsen
   it** ("the two-regime case") -- [regime()], for when you don't want to
   commit to a single elicitation.
3. **The diagnostics** -- whether pooling was actually valid, the
   fallback for when it wasn't, and how to label the pooled estimate as
   an FSATT, ATT, or ATE.

A last section collects this into a checklist for adapting the template
to your own data.

```{r setup}
library(ecem)
```

## Example data

Everything below runs on `simulate_population()`, a small data-generating
function included with the package for examples and demos -- not a
substitute for your own data, but a stand-in so this vignette is fully
self-contained and fast to run. It returns `age`, `income`, `educ`,
`region` (three categories), `noise` (unrelated to treatment or
outcome), a 0/1 treatment `D`, and an outcome `Y`. `tau0` is the true
treatment effect baked into the simulation, so you have a known answer
to check the pooled estimate against.

```{r}
set.seed(20260805)
pop <- simulate_population(N = 2000, tau0 = 3, heterogeneous = FALSE)
head(pop)
```

## Part 1: One regime -- a single elicited covariate

Suppose `age` is the covariate you want to coarsen, and you have one
substantive theory of where the breakpoints should fall: a *lifecycle*
account, with breakpoints somewhere around the end of early adulthood,
midlife, and retirement. You're not sure of the exact ages -- that
uncertainty is exactly what an elicited cutpoint captures.

Every covariate in a `cutpoint_specs` list is one of five things (see
`?regime` for the full description): a list of ranges (elicited, drawn
fresh each time), a plain numeric vector (fixed, identical every draw),
`"exact"` (match on the raw value), `NULL` (excluded from matching), or
a [regime()] object (Part 2, below). Here, `age` is elicited across
three uncertainty ranges (three breakpoints, so four age bins); `educ`
is fixed at two cutpoints the analyst is comfortable committing to;
`region` matches exactly on its three raw categories; `income` and
`noise` are excluded from matching entirely.

```{r}
cutpoint_specs_1 <- list(
  age    = list(c(22, 28), c(40, 48), c(62, 68)),  # elicited (X_E)
  educ   = c(12, 16),                              # fixed (X_F)
  region = "exact",                                # exact match
  income = NULL,                                   # excluded
  noise  = NULL                                     # excluded
)
```

`run_M_draws()` draws `M` coarsenings from this spec, matches, and
estimates the effect on each draw's matched sample -- by default via a
CEM-weighted regression of `Y` on `D` and the matching covariates (see
`?fit_effect` for the alternative, a simpler stratum mean-difference).

```{r}
set.seed(1)
draws_1 <- run_M_draws(pop, treat_var = "D", outcome_var = "Y",
                        cutpoint_specs = cutpoint_specs_1, M = 30)
draws_1
```

Each row above is one draw's own `tau_hat`; `pool_draws()` combines
them via Rubin's rules into a single point estimate and confidence
interval. Since `tau0 = 3` was the effect used to simulate this data,
`tau_bar` should land in the neighborhood of 3.

```{r}
pooled_1 <- pool_draws(draws_1)
pooled_1
```

`as.data.frame()` on the draws object gives you the same per-draw table
programmatically -- useful for plotting the spread of `tau_hat` across
draws, or feeding into your own downstream code.

```{r}
df_1 <- as.data.frame(draws_1)
str(df_1)
```

## Part 2: Two regimes -- competing theories of where to coarsen

Now suppose a colleague argues for a *generational* account instead --
coarsening `age` at cohort boundaries rather than lifecycle stages. You
don't have to pick one: [regime()] lets a covariate draw from either of
two (or more) named, competing elicited cutpoint sets, with a draw
always using one theory or the other in full -- never, say, the first
lifecycle breakpoint mixed with the last generational one.

```{r}
cutpoint_specs_2 <- list(
  age = regime(
    lifecycle    = list(c(22, 28), c(40, 48), c(62, 68)),
    generational = list(c(24, 28), c(42, 46), c(58, 62), c(76, 80))
  ),
  educ   = c(12, 16),
  region = "exact",
  income = NULL,
  noise  = NULL
)
```

By default each named regime is equally likely to be drawn (pass
`weights = ` to `regime()` to change that). Everything else about the
workflow is identical -- `run_M_draws()` doesn't need to know a
covariate has competing regimes rather than one:

```{r}
set.seed(2)
draws_2 <- run_M_draws(pop, "D", "Y", cutpoint_specs_2, M = 30)
```

`as.data.frame()` gains a `regime_age` column whenever a covariate used
[regime()], recording which theory each draw actually used:

```{r}
df_2 <- as.data.frame(draws_2)
table(df_2$regime_age)
```

```{r}
pooled_2 <- pool_draws(draws_2)
pooled_2
```

If the two theories are both defensible ways of expressing the same
underlying coarsening judgment, `tau_bar` above pools over both without
you having to choose. If you also want to see whether the two theories
give meaningfully different answers on their own, compare `tau_hat`
within each (informal -- not a formal test of whether they differ):

```{r}
tapply(df_2$tau_hat, df_2$regime_age, mean, na.rm = TRUE)
```

(When the achievable number of distinct configurations is small enough
-- e.g. because an elicited covariate has few practically distinct
observed values within its range -- `run_M_draws(..., exact_if_K_leq =)`
replaces Monte Carlo draws with exact enumeration, eliminating
simulation error outright. `demo("achievable-configuration-ceiling",
"ecem")` walks through that case; it isn't repeated here since the
draw-match-pool code above is unchanged either way.)

## Part 3: Diagnostics -- is pooling actually trustworthy?

A pooled `tau_bar` is only as trustworthy as the assumption that every
draw is estimating the same underlying quantity. `pooling_diagnostics()`
bundles the checks for that: a pre-matching flatness test on each
elicited covariate, and a bootstrap excess-variance test computed
directly on the realized draws (`n_boot` is kept small below so this
vignette runs quickly -- use several hundred for actual inference).

```{r}
set.seed(3)
diag_1 <- pooling_diagnostics(pop, draws_1, pooled = pooled_1,
                               n_boot = 100, progress = FALSE)
diag_1
```

With `heterogeneous = FALSE` in `simulate_population()`, the true effect
really is flat in `age`, so both tests should typically fail to reject,
and the printout above should recommend pooling with no caveat.

### When congeniality fails

To see the other branch of the workflow, simulate a population where the
effect genuinely does vary with the elicited covariate
(`heterogeneous = TRUE` makes `tau(age)` a real linear function of
`age`, which is exactly what the flatness/excess-variance tests are
built to catch):

```{r}
set.seed(4)
pop_het <- simulate_population(N = 2000, tau0 = 3, heterogeneous = TRUE)
draws_het <- run_M_draws(pop_het, "D", "Y", cutpoint_specs_1, M = 30)
pooled_het <- pool_draws(draws_het)

diag_het <- pooling_diagnostics(pop_het, draws_het, pooled = pooled_het,
                                 n_boot = 100, progress = FALSE)
diag_het
```

If excess-variance rejects here, `pool_draws()`'s `tau_bar` is no longer
trustworthy as a single pooled estimate, regardless of what the
flatness test says. The recommended fallback is a Simonsohn-style
existence test -- and since `pooling_diagnostics()` already bootstrapped
matches on this same `draws_het`, passing it in as `diagnostics =` reuses
that bootstrap instead of rerunning one from scratch:

```{r}
existence_test(pop_het, draws_het, diagnostics = diag_het)
```

Rejecting here is evidence that *some* nonzero effect exists somewhere
among the draws -- not validation of any single pooled point estimate.
When you land in this branch on real data, report the individual draws
descriptively (`as.data.frame(draws_het)`) rather than a pooled
`tau_bar`.

### Which estimand does the pooled estimate target?

Separately from whether pooling is *valid*, `label_diagnostics()`
addresses what the pooled estimate actually *is* an estimate of: the
feasible-sample ATT (FSATT, defined on whichever units happened to be
retained), the ATT, or the ATE. Two covariance-based diagnostics
separate these -- see `?label_diagnostics` for the full detail -- and
the print method below suggests a label based on whether each term is
large relative to the pooled estimate's own standard error (a heuristic
this package proposes, not a threshold the paper specifies).

```{r}
label_diagnostics(pop, draws_1)
```

Per the paper, treat these numbers as interpretive aids for how to
*report* `tau_bar` -- never as a correction to subtract off the point
estimate itself.

## Adapting this template to your own data

Swap `simulate_population()` for `read.csv()` (or however your data
gets loaded) and work through this checklist:

- `treat_var` needs to be a 0/1 column; `outcome_var` numeric.
- For each covariate, decide which of the five `cutpoint_specs` kinds it
  is: **elicited** (you're genuinely uncertain where to coarsen it, or
  it's the covariate a theoretical dispute is about), **fixed** (you're
  confident in a specific coarsening), `"exact"` (categorical, no
  coarsening needed), or `NULL` (excluded from matching). Elicited
  ranges should reflect real uncertainty or a real disagreement in the
  literature -- not be padded out arbitrarily to widen the prior.
- If two (or more) genuinely competing accounts exist for how a
  covariate should be coarsened, use [regime()] instead of picking one
  and hoping it's right.
- Choose `M` (or `exact_if_K_leq` if the achievable-configuration count
  is small -- check with `count_achievable_configs()`).
- Always run `pooling_diagnostics()` before reporting a pooled `tau_bar`
  from `pool_draws()`. If excess-variance rejects, fall back to
  `existence_test()` and report draws descriptively instead.
- Run `label_diagnostics()` to know whether to report the result as an
  FSATT, ATT, or ATE.

`demo(package = "ecem")` lists the package's other worked examples
(exact enumeration vs. Monte Carlo, competing regimes in more depth);
`?ecem` is the package-level overview with a map of every exported
function.
